Relict deciduous oaks show high resilience to drought despite the effect of land-use legacies

### 2019-03-28 man/ms.Rmd
update ms

In this work we used remote sensing information and dendroecological methods to evaluate the drought impacts in both canopy greenness (as a proxy to primary growth) and tree-radial growth of a Mediterranean tree species (*Quercus pyrenaica* Willd.) located in Sierra Nevada, which is considered a rear-edge of species distribution and species performance is considered to be threatened. We also assess the resilience of these forests both to several extreme drought episodes and to climate change in the long-term in the last decades.   
Our hypothesis is that these relict populations driven by historical land-use at their climatic (either altitudinal or latitudinal) rear-edge are particularly vulnerable to climate change, and hence they will show low-values of resilience after several extreme droughts. Our specifics aims were: (*i*) To quantify how two recent extreme drought events influenced primary and secondary growth of Q. pyrenaica forests in their rear edge; (*ii*) to analyze the long-term resilience of these forests to extreme drought events using time-series of radial growth; (*iii*) and to explore differences in the resilience metrics between populations located in contrasting ecological conditions within the rear edge of the distribution of this species to detect vulnerability to climate change. In addition, within the region of study we were interested to assess whether the effect of aspect and microclimatic conditions expressed in northern and southern populations of Pyrenean oak forests differ in their resistance, resilience and recovery to extreme drought events.

@@

During the last decades, the south of Iberian Peninsula suffered two of the worst recorded drought [2005 and 2012; @GarciaHerrera2007; @Trigo2013; @VicenteSerrano2014] (Table S3). However, we found a positive trend for vegetation greenness of *Q. pyrenaica* (78.9 % of the pixels showed positive trend  during the 2000 to 2016 period). This confirms previous findings findings that pointed out an increase in primary productivity for these forests [@PerezLuque2015onto], being the most productive ecosystems in this mountain region during that period [@Alcaraz2016obsnev_ndvi]. For tree-growth, we also observed positive trends in the last decades, particularly for the southern high-elevation site (CA-High, Figure 4). Similar long-term trends were described for this species along their distribution range [@GeaIzquierdo2014; @RubioCuadrado2018] and could be related to the rising of the temperatures in the last decades [@GeaIzquierdo2014]. Our pattern differs from the decline trends observed for several mediterranean and temperate tree-species located in their rear-edges [*e.g.*, *Pinus nigra*; *P. sylvestris*; @SanchezSalguero2012; @Camarero2015; *Fagus sylvatica*; @Dorado2017].

@@

As we mentioned above, we found strong declines of tree-growth during the most severe drought events occurred (1995 and 1999) (Table S3; Figure S3). This pattern was also observed for several *Pinus* species in their rear edge [@SanchezSalguero2012; @Camarero2015]. But, interestingly, we obtained the highest values of resilience for tree-growth for these two drought events (Figure 6). In addition, the analysis from a long-term perspective of the severe droughts suggests a positive relation between the tree-growth resilience and drought severity (Figures 6). Surprisingly this result contradicts our initial hypothesis in which we predict low resilience values for this oak species in their rear-edge.

Several considerations have to be taken into account to explain this result. First of all we need to explore the cumulative precipitation of the previous and posterior years of these drought events. For instance, the 1992-1994 years (pre-1995 drought) had precipitation values below the mean of the whole period (Figure S1), while 1996-1998 years (post-1995 drought) showed a wet opposite pattern. In fact, the first months of 1996 recorded the highest cumulative rainfall of the past century for our southern site (Cáñar) [@MesaTorres2009]. In addition, this argument could also explain the higher values of recovery found for 1995 (Figure 6).

@@

In addition, from a long-term perspective, we obtained a positive relation between tree-growth resilience and drought severity (Figure 6). Surprisingly, this result contradicts our initial hypothesis in which we predict low resilience values for this oak species in their rear-edge.

@@

Desde una perspectiva a largo plazo, es destacable el hecho de que las poblaciones de *Q. pyrenaica* en Sierra Nevada, aún habiendo sufrido varios eventos de sequía severa, presentan altos valores de Recovery y de Resiliencia, tal y como indican nuestros resultados (Figure 6). Las poblaciones situadas en su rear edge viven en estrechos márgenes ambientales y pequeños variaciones en las condiciones ambientales pueden provocar que las restricciones ambientales sean mas severas [@Hampe2005]. Así, frecuentemente se asume una alta vulnerabilidad a la sequía de las poblaciones situadas en su dry rear-edge [@MartinezVilalta2018], tal y como se ha visto para *Q.pyrenaica* [@GeaIzquierdo2014] y para otras especies (citas). Sin embargo, algunos estudios están demostrando que esto no es siempre así [ver por ejemplo @Cavin2017; @Granda2017], tal y como sugieren nuestros resultados, con altos valores de resiliencia para poblaciones de *Q. pyrenaica* situadas en su rear-edge.

@@

@@

The Mediterranean mountains, even at high elevations, have suffered strongly transformation of the landscapes driven by human activities throughout history [@Regato2008]. Multiple evidence from palaeoecological studies indicated an intense human impact on vegetation of Sierra Nevada since 3000 cal year BP [@Anderson2011; @JimenezMoreno2013; @GarciaAlix2017, @MesaFernandez2018]. From that moment onwards, increases on fire occurrence and both grazing and mining activities, were recorded for this mountain region, with an sharp intensification of the human activities in the last 150 years [@GarciaAlix2017; @MesaFernandez2018]. Mining, forest clearings, fuelwood and charcoal exploitations, pastoralism and also wars, have strongly impacted on the forest resources in Sierra Nevada. As a consequence of all those activities a loss of about 90% of broadleaf *Quercus* species cover from medium and low elevation occurred in this mountain region [@JimenezOlivencia1991].

@@

In Northern sites (*e.g.* San Juan), there is a sequentially distribution of the land uses along the elevational gradient [CITA CATALOGO]. Grasslands and shrublands for cattle farming are located at high elevations. Then forests formation with some croplands; and then, at lowlands, irrigated terraces with tree crops. Another activities could conditioning the natural resources uses. For instance, San Juan oak woodland is located in an area with a high concentration of mines and quarries that have been exploited intermittently throughout history (citas), with two periods of intense activity, the second half of the 19th century after the publication of detailed mineralogical reports by [@MaestreXXXX]; and during the first decades of the twentieth century until 1960, which is the last year in which there is evidence of the existence of mining activity [cita]. The mining activity in San Juan site have strongly conditioned the forest structure.

On the other hand, woodland areas of the southern slopes (*e.g.* Cáñar) are mixed with a greater percentage of croplands, even reached high elevation (mainly barley, rye and potatoes)[@Calatrava2019]. Forest resources (firewood, charcoal, acorns) were used in a continuously way, as some historical sources have pointed out [@ValbuenaCarabana2014, @JimenezSerrano2004]. In fact, in Cáñar there were an arraigated tradition of charcoal extraction since fifteenth century [@JimenezSerrano2004] to late twenty century [@MesaTorres2009; @Bonet2014_conama].

@@

The release event observed for SJ site during 1995-2000 was lower than the occurred at 1940, but affecting more trees (Figures 5, S4). The revision of forests practices carried out in this area in the last 30 years did not reveal any clearing or cutting event for those years. Two non-mutually exclusive ways could explain the tree-growth release observed. Firstly it could be related with a natural drought-induced mortality event after 1995, as was reported for Mediterranean tree species [*e.g.* @Penuelas2000; @Lloret2004; @Gentilesca2017]. On the other hand, the strong positive correlations of SPEI with tree-growth obtained for this site (Figure 8) indicates a high sensitivity to water availability of this slightly more xeric site (Table 1) [@GeaIzquierdo2014]. Previous works reported a sharp decline (or non-production) in latewood production during extreme drought events [@Corcuera2006]. Since latewood are less vulnerable to embolism than earlywood vessels [@Corcuera2006], the reduction or even the non-production of former could negatively affect to tree-growth and also enhanced the mortality, particularly for drier sites [@Corcuera2006; @GeaIzquierdo2014].

>>>>

In this work we used remote sensing information and dendroecological methods to evaluate the drought impacts in both canopy greenness (as a proxy to primary growth) and tree-radial growth of a Mediterranean tree species (*Quercus pyrenaica* Willd.) located in Sierra Nevada, which is considered a rear-edge of species distribution and species performance is considered to be threatened. We also assess the resilience of these forests both to several extreme drought episodes and to climate change in the long-term in the last decades.   
Our hypothesis is that these relict populations driven by historical land-use at their climatic (either altitudinal or latitudinal) rear-edge are particularly vulnerable to climate change, and hence they will show low-values of resilience after several extreme droughts. Our specifics aims were: (*i*) To quantify how two recent extreme drought events influenced primary and secondary growth of *Q. pyrenaica* forests in their rear edge; (*ii*) to analyze the long-term resilience of these forests to extreme drought events using time-series of radial growth; (*iii*) and to explore differences in the resilience metrics between populations located in contrasting ecological conditions within the rear edge of the distribution of this species to detect vulnerability to climate change. In addition, within the region of study we were interested to assess whether the effect of aspect and microclimatic conditions expressed in northern and southern populations of Pyrenean oak forests differ in their resistance, resilience and recovery to extreme drought events.

@@

During the last decades, the south of Iberian Peninsula suffered two of the worst recorded drought [2005 and 2012; @GarciaHerrera2007; @Trigo2013; @VicenteSerrano2014] (Table S3). However, we found a positive trend for vegetation greenness of *Q. pyrenaica* (78.9 % of the pixels showed positive trend  during the 2000 to 2016 period). This confirms previous findings that pointed out an increase in primary productivity for these forests [@PerezLuque2015onto], being the most productive ecosystems in this mountain region during that period [@Alcaraz2016obsnev_ndvi]. For tree-growth, we also observed positive trends in the last decades, particularly for the southern high-elevation site (CA-High, Figure 4). Similar long-term trends were described for this species along their distribution range [@GeaIzquierdo2014; @RubioCuadrado2018] and could be related to the rising of the temperatures in the last decades [@GeaIzquierdo2014]. Our pattern differs from the decline trends observed for several mediterranean and temperate tree-species located in their rear-edges [*e.g.*, *Pinus nigra*; *P. sylvestris*; @SanchezSalguero2012; @Camarero2015; *Fagus sylvatica*; @Dorado2017].

@@

As we mentioned previously, we found strong declines of tree-growth during the most severe drought events occurred (1995 and 1999) (Table S3; Figure S3). This pattern was also observed for other species in their rear edge [@SanchezSalguero2012; @Camarero2015]. But, interestingly, we obtained the highest values of resilience for tree-growth for these two drought events (Figure 6). In addition, the analysis from a long-term perspective of the severe droughts suggests a positive relation between the tree-growth resilience and drought severity (Figures 6). Surprisingly this result seems contradict our initial hypothesis in which we predict low resilience values for this oak species in their rear-edge after extreme drought events.

Rear-edge populations live in environmental narrow margins, and small variations in environmental conditions can increase the vulnerability of the species [@Hampe2005]. For instance, higher vulnerability to drought is ofted assumed for rear-edge populations [@MartinezVilalta2018], however some studies showed different findings [@Cavin2017; @Granda2017], as our results suggest, with high values of resilience for populations of * Q. pyrenaica * located in its rear-edge.

This would be related with the definition of the marginal population. Algunos autores han apuntando que cuando se estudian poblaciones del rear-edge, hay que poner atención a la forma en la que se define la marginalidad, esto es, si se define atendiendo a criterios geográficos, climáticos, o según otros factores ecológicos [@MartinezVilalta2018]. En este sentido, los altos valores de resiliencia a los eventos de sequía que hemos observado, podrían sugerir que las poblaciones de *Q. pyrenaica* en Sierra Nevada están situados en un rear-edge geográfico, pero no climático. Esto último se ve reforzado por el hecho de que Sierra Nevada is considered a glacial refugia for deciduous *Quercus* species [@Brewer2002; @Olalde2002; @RodriguezSanchez2010]; además las poblaciones de *Q. pyrenaica* en Sierra Nevada presentan una alta resiliencia genética [@Valbuena2013; @Valbuena2017]

Several considerations have to be taken into account to explain this result. Firstly it would be interesting to explore the cumulative precipitation of the previous and posterior years of these drought events. For instance, the 1992-1994 years (pre-1995 drought) had precipitation values below the mean of the whole period (Figure S1), while 1996-1998 years (post-1995 drought) showed a wet opposite pattern. In fact, the first months of 1996 recorded the highest cumulative rainfall of the past century for our southern site (Cáñar) [@MesaTorres2009]. In addition, this argument could also explain the higher values of recovery found for 1995 (Figure 6).

A plausible explanation could be

@@

@@

Desde una perspectiva a largo plazo, es destacable el hecho de que las poblaciones de *Q. pyrenaica* en Sierra Nevada, aún habiendo sufrido varios eventos de sequía severa, presentan altos valores de Recovery y de Resiliencia, tal y como indican nuestros resultados (Figure 6).

@@

Las poblaciones situadas en su rear edge viven en estrechos márgenes ambientales y pequeños variaciones en las condiciones ambientales pueden provocar que las restricciones ambientales sean mas severas [@Hampe2005]. Así, frecuentemente se asume una alta vulnerabilidad a la sequía de las poblaciones situadas en su dry rear-edge [@MartinezVilalta2018], tal y como se ha visto para *Q.pyrenaica* [@GeaIzquierdo2014] y para otras especies (citas). Sin embargo, algunos estudios están demostrando que esto no es siempre así [ver por ejemplo @Cavin2017; @Granda2017], tal y como sugieren nuestros resultados, con altos valores de resiliencia para poblaciones de *Q. pyrenaica* situadas en su rear-edge.

@@

The Mediterranean mountains, even at high elevations, have suffered strongly transformation of the landscapes driven by human activities throughout history [@Regato2008]. Multiple evidence from palaeoecological studies indicated an intense human impact on vegetation of Sierra Nevada since 3000 cal year BP [@Anderson2011; @JimenezMoreno2013; @GarciaAlix2017; @MesaFernandez2018]. From that moment onwards, increases on fire occurrence and both grazing and mining activities, were recorded for this mountain region, with an sharp intensification of the human activities in the last 150 years [@GarciaAlix2017; @MesaFernandez2018]. Mining, forest clearings, fuelwood and charcoal exploitations, pastoralism and also wars, have strongly impacted on the forest resources in Sierra Nevada. As a consequence of all those activities a loss of about 90% of broadleaf *Quercus* species cover from medium and low elevation occurred in this mountain region [@JimenezOlivencia1991].

@@

In Northern sites (e.g. San Juan), there is a sequentially distribution of the land uses along the elevational gradient [CITA CATALOGO]. Grasslands and shrublands for cattle farming are located at high elevations. Then forests formation with some croplands; and then, at lowlands, irrigated terraces with tree crops. Another activities could conditioning the natural resources uses. For instance, the mining activity in San Juan site have strongly conditioned the forest structure. This oak woodland is located in an area with a high concentration of mines and quarries that have been exploited intermittently throughout history (citas).  Historical documents indicated two periods of intense mining activity: the second half of the 19th century after the publication of detailed mineralogical reports by [@MaestreXXXX]; and during the first decades of the twentieth century until 1960, which is the last year in which there is evidence of the existence of mining activity [cita].

On the other hand, woodland areas of the southern slopes (Cáñar) are mixed with a greater percentage of croplands, even reached high elevation (mainly barley, rye and potatoes) [@Calatrava2019]. Forest resources, like firewood, charcoal, acorns, have been continuously exploited in this site through history [@ValbuenaCarabana2014, @JimenezSerrano2004]. In Cáñar there were an arraigated tradition of charcoal extraction since fifteenth century [@JimenezSerrano2004] up to middle of the 20th century. From this moment to the present, there were a sharp decrease of the wood extraction, mainly due to rural abandonment [@MesaTorres2009; @Bonet2014_conama].

@@

The other release event observed for SJ site during 1995-2000 was lower than the occurred at 1940, but affecting more trees (Figures 5, S4). We revised the forest practices carried out in this area in the last 30 years [CITA], but we did not find any clearing or cutting event during 1995 - 2000 period. Two non-mutually exclusive ways could explain the tree-growth release observed. Firstly it could be related with a natural drought-induced mortality event after 1995, as was reported for Mediterranean tree species [*e.g.* @Penuelas2000; @Lloret2004; @Gentilesca2017]. On the other hand, we obtained strong positive correlations of SPEI with tree-growth for this site (Figure 8), which suggests a high sensitivity to water availability of this slightly more xeric site (Table 1) [@GeaIzquierdo2014]. Previous works reported a sharp decline (or non-production) in latewood production during extreme drought events [@Corcuera2006]. Since latewood are less vulnerable to embolism than earlywood vessels [@Corcuera2006], the reduction or even the non-production of former could negatively affect to tree-growth and also enhanced the mortality, particularly for drier sites [@Corcuera2006; @GeaIzquierdo2014].
